---
title: "Data Cleaning in R"
author: "Sayorn Chin"
date: "06/09/2021"
output:
  github_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---


### Load required libraries
```{r,warning=FALSE,message=FALSE}
load_libraries <- function(){
  library(readr)
  library(tidyverse)
  library(dplyr)
  library(lubridate)
  library(broom)
  library(ggplot2)
  print('The libraries have been loaded.')
}
load_libraries()
```

### Section I

#### Load bike dataset
```{r, warning=FALSE, message=FALSE}
setwd('/Users/schinlfc/data-science-R/data-cleaning-R/data/')
bike = read_csv('bike_lanes.csv')
```

##### 1. Inspect the entire dataset
```{r, warning=FALSE, message=FALSE}
str(bike) 
summary(bike) 
names(bike) 
ncol(bike) 
nrow(bike) 
sum(complete.cases(bike))
sum(!complete.cases(bike)) 
```

##### 2. Create a data set called `namat` which is equal to `is.na(bike)`. What is the class of `namat`?  Run rowSums and colSums on `namat.` These represent the number of missing values in the rows and columns of `bike`. Don't print `rowSums`, but do a table of the `rowSums`. Try `nabike = bike %>% mutate_all(is.na)`. Then try `nabike %>% summarise_all(sum)`.  What do you get?
```{r, warning=FALSE, message=FALSE}
namat <- is.na(bike)
class(namat)
table(rowSums(namat))
colSums(namat)
nabike <- bike %>% mutate_all(is.na)
print(nabike)
print(nabike %>% summarise_all(sum))
```

##### 3. Filter rows of bike that are NOT missing the `route` variable, assign this to the object `have_route.`  Do a table of the `subType` using `table`, including the missing `subType`s  Get the same frequency distribution using `group_by(subType)` and `tally()`
```{r, warning=FALSE, message=FALSE}
have_route <- bike %>% filter(!is.na(route))
print(have_route)
table(bike$subType, useNA='always')
print(have_route %>% group_by(subType) %>% tally())
print(have_route %>%  count(subType))
print(have_route %>% group_by(subType) %>% summarize(n_obs = n()))
print(tally(group_by(have_route, subType)))
have_route = group_by(have_route, subType)
print(tally(have_route))
```

##### 4. Filter rows of bike that have the type `SIDEPATH` or `BIKE LANE` using `%in%`.  Call it `side_bike.` Confirm this gives you the same number of results using the `|` and `==`.
```{r, warning=FALSE, message=FALSE}
side_bike <- bike %>% filter(type %in% c('SIDEPATH', 'BIKE LANE'))
print(side_bike)
side_bike <- bike %>% filter(type == 'SIDEPATH' | type == 'BIKE LANE')
print(side_bike)
```

##### 5. Do a cross tabulation of the bike `type` and the number of lanes. Call it `tab`.  Do a `prop.table` on the rows and columns margins. Try `as.data.frame(tab)` or `broom::tidy(tab)`
```{r, warning=FALSE, message=FALSE}
tab <- table(bike_type = bike$type, number_of_lanes = bike$numLanes)
print(tab)
prop.table(tab, 1)
prop.table(tab, 2)
print(as.data.frame(tab))
print(broom::tidy(tab))
```

### Section II

#### Load tax dataset
```{r, warning=FALSE, message=FALSE}
setwd('/Users/schinlfc/data-science-R/data-cleaning-R/data/')
tax <- read_csv('real_property_taxes.csv')
```

##### 6. How many addresses pay property taxes?
```{r, warning=FALSE, message=FALSE}
dim(tax)
nrow(tax)
length(tax$PropertyID)
sum(is.na(tax$CityTax))
sum(!is.na(tax$CityTax))
```

##### 7. What is the total city and state tax paid?  
You need to remove the `$` from the `CityTax` variable
then you need to make it numeric. Try `str_replace`, but remember
$ is "special" and you need `fixed()` around it.

###### 7.1 City
```{r, warning=FALSE, message=FALSE}
head(tax$CityTax) 
tax <- tax %>% 
  mutate(CityTax = str_replace(CityTax, fixed("$"), ""),
  CityTax = as.numeric(CityTax))
head(tax$CityTax)
tax$CityTax <- str_replace(tax$CityTax, fixed("$"), "")
tax$CityTax <- as.numeric(tax$CityTax)
head(tax$CityTax)
sum(is.na(tax$CityTax))
options(digits=12) # so no rounding
sum(tax$CityTax, na.rm = TRUE)
sum(tax$CityTax, na.rm = TRUE) / 1e6
```

###### 7.2 State
```{r, warning=FALSE, message=FALSE}
options(digits=12) # so no rounding
head(tax$StateTax)
tax <- tax %>% mutate(StateTax = parse_number(StateTax))
head(tax$StateTax)
sum(is.na(tax$StateTax))
sum(tax$StateTax, na.rm = TRUE)
sum(tax$StateTax, na.rm = TRUE) / 1e6
```

##### 8. Using `table()` or `group_by` and `summarize(n())` or `tally()`
###### 8.1 how many observations/properties are in each Ward?
```{r, warning=FALSE, message=FALSE}
table(tax$Ward)
ward_table <- tax %>% group_by(Ward) %>% tally()
print(ward_table)
ward_table <- tax %>% group_by(Ward) %>%
  summarise(number_of_observations = n())
print(ward_table)
```

###### 8.2 what is the mean state tax per Ward? use `group_by` and `summarize
```{r, warning=FALSE, message=FALSE}
mean_statetax <- tax %>% group_by(Ward) %>%
  summarise(mean_statetax = mean(StateTax, na.rm = TRUE))
print(mean_statetax)
```

###### 8.3 what is the maximum amount still due in each Ward?  different summarization (`max`)
```{r, warning=FALSE, message=FALSE}
tax <- tax %>% mutate(AmountDue = str_replace(
  AmountDue, fixed("$"), ""), AmountDue = as.numeric(AmountDue))
head(tax$AmountDue)

tax$AmountDue = tax$AmountDue %>% 
  str_replace(fixed("$"), "") %>%
  as.numeric
head(tax$AmountDue)

tax = tax %>% mutate(AmountDue = as.numeric(str_replace(
    AmountDue, fixed("$"), "")))
head(tax$AmountDue)

max_due <- tax %>% group_by(Ward) %>% 
  summarise(max_amount_due = max(AmountDue, na.rm = TRUE))
print(max_due)
```

###### 8.4 what is the 75th percentile of city and state tax paid by Ward? (`quantile`)
```{r, warning=FALSE, message=FALSE}
percentile_city <- tax %>% group_by(Ward) %>% 
  summarise(percentile_city = quantile(
    CityTax, prob = 0.75, na.rm = TRUE))
print(percentile_city)

percential_state <- tax %>% group_by(Ward) %>%
  summarise(percential_state = quantile(
    StateTax, prob = 0.75, na.rm = TRUE))
print(percential_state)

percentile_city_state <- tax %>% 
  group_by(Ward) %>% summarise(
    percentile_city = quantile(CityTax, prob = 0.75, na.rm =TRUE),
    percentile_state = quantile(StateTax, prob = 0.75, na.rm =TRUE))
print(percentile_city_state)
```

##### 9. Make boxplots using showing `cityTax` (`y` -variable) by whether the property	is a principal residence (`x`) or not.
```{r, warning=FALSE, message=FALSE}
tax <- tax %>% mutate(CityTax = as.numeric(str_replace(CityTax, fixed("$"), "")),
                      ResCode = str_trim(ResCode))

ggplot(tax, aes(x=ResCode, y=log10(CityTax+1), color=ResCode)) +
  geom_boxplot() +
  labs(title="City tax by residential code") +
  xlab("Residential code") +
  ylab("City tax")
```

##### 10. Subset the data to only retain those houses that are principal residences. Which command subsets rows? Filter or select?

###### 10.1 How many such houses are there?
```{r, warning=FALSE, message=FALSE}
pres = tax %>% filter( ResCode %in% "PRINCIPAL RESIDENCE")
nrow(pres)
pres = tax %>% filter( ResCode == "PRINCIPAL RESIDENCE")
nrow(pres)
```

###### 10.2 Describe the distribution of property taxes on these residences.
```{r, warning=FALSE, message=FALSE}
pres <- pres %>% mutate(StateTax = as.numeric(str_replace(StateTax, fixed("$"), "")))

ggplot(pres, aes(x=log(CityTax))) + 
  geom_histogram(binwidth=0.3, color="darkblue",
  fill="lightblue")

ggplot(pres, aes(x=log(CityTax)))+
  geom_density(color="darkblue", fill="lightblue")

ggplot(pres, aes(x=log(StateTax))) + 
  geom_histogram(binwidth=0.3, color="darkblue",
  fill="lightblue")

ggplot(pres, aes(x=log(StateTax)))+
  geom_density(color="darkblue", fill="lightblue")
```

### Section III

#### Load salary dataset
```{r, warning=FALSE, message=FALSE}
setwd('/Users/schinlfc/data-science-R/data-cleaning-R/data/')
sal <- read_csv('baltimore_city_employee_salaries_FY2015.csv')
```

##### 11. Make an object called health.sal using the salaries data set, with only agencies of those with `"fire"` (or any forms), if any, in the name remember `fixed( ignore_case = TRUE)` will ignore cases.
```{r, warning=FALSE, message=FALSE}
health.sal <- sal %>% filter(str_detect(Agency, fixed("fire", ignore_case = TRUE)))
print(head(health.sal))
```

##### 12. Make a data set called `trans` which contains only agencies that contain "TRANS".
```{r, warning=FALSE, message=FALSE}
trans <- sal %>% filter(str_detect(Agency, "TRANS"))
print(head(trans))
```

##### 13. What is/are the profession(s) of people who have `"abra"` in their name for Baltimore's Salaries?  Case should be ignored. 
```{r, warning=FALSE, message=FALSE}
print(sal %>% 
  filter(str_detect(name, fixed("abra", ignore_case = TRUE))) %>% select(name, JobTitle))
```

##### 14. What is the distribution of annual salaries look like? (use `hist`) What is the IQR? Hint: first convert to numeric. Try `str_replace`, but remember`$` is "special" and you need `fixed()` around it.
```{r, warning=FALSE, message=FALSE}
sal <- sal %>% 
  mutate(AnnualSalary = as.numeric(
  str_replace(AnnualSalary, fixed("$"), "")))
print(head(sal))

ggplot(sal, aes(x=log(AnnualSalary))) + 
  geom_histogram(binwidth=0.3, color="darkblue",
  fill="lightblue")
```